---
title: "Bayesian latent class models for prevalence estimation and diagnostic test evaluation"
author: "Simon Dufour and Juan Carlos Arango Sabogal"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography:
- book.bib
- packages.bib
biblio-style: apalike
link-citations: yes
description: Chapitre 12
output: html_document
---


```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```


# Exercise 5 - Accuracy varying with covariate
## Questions
  
Open up the `R` project for Exercise 5 (i.e., the R project file named *Exercise 5.Rproj*).   
  
In the Exercise 5 folder, you will find partially completed `R` scripts. To answer the questions, try to complete the missing parts (they are highlighted by a **#TO COMPLETE#** comment). We also provided complete `R` scripts, but try to work it out on your own first! 
  
Some anecdotic reports suggest that expression of PAG may vary as function of number of days of pregnancy. We could, therefore, hypothesize that PAG's Se and Sp may vary if the test is conducted early (≤ 35 days post-breeding) compared to later (>35 days post-breeding). When cross-tabulating the studies results for the PAG and US tests as function of days since breeding, we get:  
  
**Table.** Cross-classified results from a PAG and US test as function of number of days since breeding.  
  
| **≤ 35 days** |         |         |   | **>35 days** |         |         |
|:-------------:|:-------:|:-------:|:-:|:------------:|:-------:|:-------:|
|               | **US+** | **US-** |   |              | **US+** | **US-** |
|    **PAG+**   |   143   |    11   |   |   **PAG+**   |   112   |    10   |
|    **PAG-**   |    2    |   136   |   |   **PAG-**   |    0    |    82   |
  
**1.Start from the partially completed *Question 1.R* script located in Exercise 5 folder.** Build a model for two diagnostic tests, one single population, but with Se and Sp of the PAG test varying as function of days since breeding. Use the [Romano et al. (2006)](https://www.sciencedirect.com/science/article/pii/S0093691X06001543?via%3Dihub) results, to inform the priors for the Se and Sp of a US exam (Se: 0.90, 95CI: 0.85, 0.95; Sp: 0.95, 95CI: 0.90, 0.97). Assume conditional independence between tests. How many degrees of freedom do you have and how many unknown parameters do you need to estimate? Based on your results, would you conclude that PAG accuracy is influenced by number of days since breeding?   
  
  
## Answers
**1.** How many degrees of freedom do you have and how many unknown parameters do you need to estimate? Based on your results, would you conclude that PAG accuracy is influenced by number of days since breeding?   
  
**Answer:**  We have 6 degrees of freedom and 8 unknown parameters to estimate (one Se and one Sp for US, 2 Se and 2 Sp for PAG, and two prevalence). We need two informative priors to make the LCM identifiable.   
  
```{r message=FALSE, warning=FALSE, include=FALSE}

# Generate data and labels ------------------------------------------------
#n is of the form : (TestA pos and TestB pos), (TestA pos and TestB neg), (TestA neg and TestB pos), then (TestA neg and TestB neg)
datalist <- list(Z0=c(143, 2, 11, 136),
                 Z1=c(112, 0, 10, 82)
                )

#I could first create labels for TestA and TestB
TestA <- "US"
TestB <- "PAG"

#I could create labels for Z=0 and Z=1
Z0 <- "28_35days"
Z1 <- "36_45days"

library(epiR) 

# US sensitivity ----------------------------------------------------------
# Sensitivity of US:  Mode=0.90, and we are 97.5% sure >0.85 
Se.US <- epi.betabuster(mode=0.90, conf=0.975, imsure="greater than", x=0.85)  

# US specificity ----------------------------------------------------------
# Specificity of US:  Mode=0.95, and we are 97.5% sure >0.90 
Sp.US <- epi.betabuster(mode=0.95, conf=0.975, imsure="greater than", x=0.90)  

#Provide information for the prior distributions (all beta distributions) for the unknown parameters 
Prev0.shapea <- 1         #a shape parameter for Prev in Z=0    
Prev0.shapeb <- 1         #b shape parameter for Prev in Z=0

Prev1.shapea <- 1         #a shape parameter for Prev in Z=1    
Prev1.shapeb <- 1         #b shape parameter for Prev in Z=1

Se.TestA.shapea <- Se.US$shape1     #a shape parameter for Se of TestA
Se.TestA.shapeb <- Se.US$shape2     #b shape parameter for Se of TestA
Sp.TestA.shapea <- Sp.US$shape1     #a shape parameter for Sp of TestA
Sp.TestA.shapeb <- Sp.US$shape2     #b shape parameter for Sp of TestA

Se.TestB.Z0.shapea <- 1     #a shape parameter for Se of TestB in Z=0
Se.TestB.Z0.shapeb <- 1     #b shape parameter for Se of TestB in Z=0
Sp.TestB.Z0.shapea <- 1     #a shape parameter for Sp of TestB in Z=0
Sp.TestB.Z0.shapeb <- 1     #b shape parameter for Sp of TestB in Z=0

Se.TestB.Z1.shapea <- 1     #a shape parameter for Se of TestB in Z=1
Se.TestB.Z1.shapeb <- 1     #b shape parameter for Se of TestB in Z=1
Sp.TestB.Z1.shapea <- 1     #a shape parameter for Sp of TestB in Z=1
Sp.TestB.Z1.shapeb <- 1     #b shape parameter for Sp of TestB in Z=1

#I will also need the number of individuals tested in each populations (n0 and n1)
n <- sapply(datalist, sum)
n0 <- n[1]
n1 <- n[2]


# Create the JAGS text file -----------------------------------------------
#Create the JAGS text file
model_2tests_covariate <- paste0("model{

#=== LIKELIHOOD	===#
  #For Z=0
  Z0[1:4] ~ dmulti(p0[1:4], ",n0,")
  p0[1] <- Prev_", Z0,"*Se_", TestA, "*Se_", TestB, "_", Z0," + (1-Prev_", Z0,")*(1-Sp_", TestA, ")*(1-Sp_", TestB, "_", Z0,")
  p0[2] <- Prev_", Z0,"*Se_", TestA, "*(1-Se_", TestB, "_", Z0,") + (1-Prev_", Z0,")*(1-Sp_", TestA, ")*Sp_", TestB, "_", Z0,"
  p0[3] <- Prev_", Z0,"*(1-Se_", TestA, ")*Se_", TestB, "_", Z0," + (1-Prev_", Z0,")*Sp_", TestA, "*(1-Sp_", TestB, "_", Z0,")
  p0[4] <- Prev_", Z0,"*(1-Se_", TestA, ")*(1-Se_", TestB, "_", Z0,") + (1-Prev_", Z0,")*Sp_", TestA, "*Sp_", TestB, "_", Z0,"
  
  #For Z=1
  Z1[1:4] ~ dmulti(p1[1:4], ",n1,")
  p1[1] <- Prev_", Z1,"*Se_", TestA, "*Se_", TestB, "_", Z1," + (1-Prev_", Z1,")*(1-Sp_", TestA, ")*(1-Sp_", TestB, "_", Z1,")
  p1[2] <- Prev_", Z1,"*Se_", TestA, "*(1-Se_", TestB, "_", Z1,") + (1-Prev_", Z1,")*(1-Sp_", TestA, ")*Sp_", TestB, "_", Z1,"
  p1[3] <- Prev_", Z1,"*(1-Se_", TestA, ")*Se_", TestB, "_", Z1," + (1-Prev_", Z1,")*Sp_", TestA, "*(1-Sp_", TestB, "_", Z1,")
  p1[4] <- Prev_", Z1,"*(1-Se_", TestA, ")*(1-Se_", TestB, "_", Z1,") + (1-Prev_", Z1,")*Sp_", TestA, "*Sp_", TestB, "_", Z1,"
   

#=== PRIOR	===#

  Prev_", Z0," ~ dbeta(",Prev0.shapea,", ",Prev0.shapeb,") 	## Prior for Prev in Z=0
  Prev_", Z1," ~ dbeta(",Prev1.shapea,", ",Prev1.shapeb,") 	## Prior for Prev in Z=1
  Se_", TestA, " ~ dbeta(",Se.TestA.shapea,", ",Se.TestA.shapeb,") 	## Prior for Se of Test A
  Sp_", TestA, " ~ dbeta(",Sp.TestA.shapea,", ",Sp.TestA.shapeb,") 	## Prior for Sp of Test A
  Se_", TestB, "_", Z0, " ~ dbeta(",Se.TestB.Z0.shapea,", ",Se.TestB.Z0.shapeb,") 	## Prior for Se of Test B in Z=0
  Sp_", TestB, "_", Z0, " ~ dbeta(",Sp.TestB.Z0.shapea,", ",Sp.TestB.Z0.shapeb,") 	## Prior for Sp of Test B in Z=0
  Se_", TestB, "_", Z1, " ~ dbeta(",Se.TestB.Z1.shapea,", ",Se.TestB.Z1.shapeb,") 	## Prior for Se of Test B in Z=1
  Sp_", TestB, "_", Z1, " ~ dbeta(",Sp.TestB.Z1.shapea,", ",Sp.TestB.Z1.shapeb,") 	## Prior for Sp of Test B in Z=1
  
}")

#write as a text (.txt) file
write.table(model_2tests_covariate, 
            file="model_2tests_covariate.txt", 
            quote=FALSE, 
            sep="", 
            row.names=FALSE,
            col.names=FALSE)


# Generate initial values -------------------------------------------------
inits <- list(list(Prev_28_35days=0.50,
                   Prev_36_45days=0.50,
                   Se_US=0.90,
                   Sp_US=0.90,
                   Se_PAG_28_35days=0.90,
                   Sp_PAG_28_35days=0.90,
                   Se_PAG_36_45days=0.90,
                   Sp_PAG_36_45days=0.90),
              
              list(Prev_28_35days=0.60,
                   Prev_36_45days=0.60,
                   Se_US=0.80,
                   Sp_US=0.80,
                   Se_PAG_28_35days=0.80,
                   Sp_PAG_28_35days=0.80,
                   Se_PAG_36_45days=0.80,
                   Sp_PAG_36_45days=0.80),
              
              list(Prev_28_35days=0.40,
                   Prev_36_45days=0.40,
                   Se_US=0.70,
                   Sp_US=0.70,
                   Se_PAG_28_35days=0.70,
                   Sp_PAG_28_35days=0.70,
                   Se_PAG_36_45days=0.70,
                   Sp_PAG_36_45days=0.70)
)


# Run the model -----------------------------------------------------------
library(R2jags)
library(coda)

#Run the Bayesian model
bug.out <- jags(data=datalist,                             
                model.file="model_2tests_covariate.txt",     
                parameters.to.save=c("Prev_28_35days", "Prev_36_45days","Se_US", "Sp_US", "Se_PAG_28_35days", "Sp_PAG_28_35days", "Se_PAG_36_45days", "Sp_PAG_36_45days"),               
                n.chains=3,                                 
                inits=inits,                                
                n.iter=21000,                                
                n.burnin=1000,                              
                n.thin=1,                                   
                DIC=FALSE)

# Check results -----------------------------------------------------------
print(bug.out, digits.summary=3) 
```
The results from this LCM are below.   
  
**Table.** Results for the LCM allowing for varying accuracy of PAG as function of days since breeding.  
  
|       **Parameter**       | **Median** |   **95CI**   |
|:-------------------------:|:----------:|:------------:|
| **Prevalence 28-35 days** |    0.521   | 0.462, 0.580 |
| **Prevalence 36-45 days** |    0.584   | 0.512, 0.653 |
|         **Se US**         |    0.928   | 0.895, 0.956 |
|         **Sp US**         |    0.977   | 0.957, 0.990 |
|   **Se PAG 28-35 days**   |    0.991   |  0.963, 1.00 |
|   **Se PAG 36-45 days**   |    0.994   |  0.966, 1.00 |
|   **Sp PAG 28-35 days**   |    0.980   | 0.927, 0.999 |
|   **Sp PAG 36-45 days**   |    0.962   | 0.879, 0.998 |  
  
There was little difference between the 28-35 days since breeding *vs.* 36-45 days since breeding accuracy parameters. Below, we have plotted the two PAG Sp posterior distributions to illustrate how similar they are. However, since we are not in a Frequentist framework, "testing" whether the 28-35 days *vs.* 36-45 days accuracy parameters are statistically different is not straightforward. Note that Bayesian methods can be used to compute the probability that a given parameter is larger (or smaller) than another parameter, or than an estimate previously reported in the literature [(see Arango-Sabogal et al., 2021](https://doi.org/10.1016/j.prevetmed.2021.105510) for an applied example.    
  
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Posterior distributions for PAG specificity at 28-35 days (black) and 36-45 days (red) post-breeding."}
library(dplyr)
a <- bug.out$BUGSoutput$sims.list %>%
  as.data.frame()

library(ggplot2)
P <-  ggplot(a, aes(x=Sp_PAG_28_35days))+
  geom_density(color="black") + 
    theme_bw()+
    xlim(0, 1.01)+
    xlab("Specificity")

P <- P + geom_density(aes(x=Sp_PAG_36_45days), color="red")

P

```

